The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems

### 2011-10-19 shortRange.inhomo/writeDown/main.tex
revise the confusing notation in Fourier transform

The naive calculation of the convolutions in the error estimate \eqref{eqn:tmp5} requires
a computational cost of $\mathcal O(N^2)$. To estimate the force error
on the fly, the computational cost of the estimate should be reduced to
at least $\mathcal O(N\log N)$.  This is achieved by  using the fast Fourier transform. We
assume the simulation box is uniformly divided in to small bins.  The
positions of these bins are
\begin{align}
  \v r_{i_1, i_2, i_3} & =
  \frac{i_1}{K_1}\v a_1 + 
  \frac{i_2}{K_2}\v a_2 + 
  \frac{i_3}{K_3}\v a_3,\qquad 0\leq i_\alpha< K_\alpha
\end{align}
where $\v a_\alpha$ are box vectors, and $K_\alpha$ is the number of
division on each direction. In the reciprocal space, a corresponding
lattice is set up:
\begin{align}
  \v k_{k_1, k_2, k_3} & =
  k_1\v a_1^\ast +
  k_2\v a_2^\ast +
  k_3\v a_3^\ast, \qquad 0\leq k_\alpha<K_\alpha
\end{align}
where $\v a_\alpha^\ast$ are reciprocal box vectors defined by $\v
a_\alpha\cdot\v a_\beta^\ast = \delta_{\alpha\beta}$.  In each bin the
error and the particles number density are assumed to be constant,
then, a Fourier expansion can be obtained:
\begin{align}\nonumber
  \mathcal E^2_{\textrm{homo}}(\v r_{i_1,i_2,i_3})
  & =
  \frac1V\sum_{k_1=-\infty}^\infty\sum_{k_2=-\infty}^\infty\sum_{k_3=-\infty}^\infty
  [{\mathcal E}^2_{\textrm{homo}}]^{\wedge}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg] \\\label{eqn:tmp13}
  & =
  \frac1V\sum_{k_1=0}^{K_1}\sum_{k_2=0}^{K_2}\sum_{k_3=0}^{K_3}
  \hat{K}_{\textrm{homo}}(\v k'_{k_1,k_2,k_3})\:
  \hat{\rho}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
Here we denote the convolution kernel $\vert\v f^c(\v r)\vert^2$ by
$K_{\textrm{homo}}(\v r)$.  The second equation in \eqref{eqn:tmp13}
holds due to the fact: the Fourier transform of the convolution
$K_{\textrm{homo}}\ast\rho$ is equal to the multiplication of Fourier transforms
$\hat{K}_{\textrm{homo}}$ and $\hat\rho$, which are given by
\begin{align}\label{eqn:f-kernel-homo}
  \hat K_{\textrm{homo}}(\v k)
  &=
  \int_{\mathbb R^3}K_{\textrm{homo}}(\v r)e^{-2\pi i\v k\cdot\v r}\,\d d\v r,
\end{align}
and
\begin{align}
  \hat\rho(\v k_{k_1,k_2,k_3})
  & =
  \frac{V}{K_1K_2K_3}
  \sum_{i_1=0}^{K_1}\sum_{i_2=0}^{K_2}\sum_{i_2=0}^{K_2}
  \rho(\v r_{i_1,i_2,i_3})
  \exp\bigg[
  -2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
Note that in Eqn. \eqref{eqn:tmp13}, the prime on $\v k_{k_1,k_2,k_3}$ means
that the periodic image $k_\alpha - K_\alpha$ should use instead of
$k_\alpha$ when $k_\alpha \geq K_\alpha/2$.
In Eqn. \eqref{eqn:f-kernel-homo}
the Fourier transform is performed on $K_{\textrm{homo}}(\v r)$ rather than
$K_{\textrm{homo}}(\v r - \v r')$.
Most of the short-range
interactions are isotropic, so  $K_{\textrm{homo}}(\v r) =
K_{\textrm{homo}}(r)$, then the three-dimensional integral in
Eqn. \eqref{eqn:f-kernel-homo} can be simplified to a one-dimensional
integral:
\begin{align}\label{eqn:kernel-inte}
  \hat{K}_{\textrm{homo}}(\v k)
  =
  % \int_{\mathbb R^3}K_{\textrm{homo}}(r)e^{-2\pi i\v k\cdot\v r}\d d\v r
  % =
  \int_{r_c}^\infty \frac{2 r}k\, K_{\textrm{homo}}(r) \sin(2\pi kr)\,\d dr,
\end{align}
which can be calculated by numerical
integration from $r_c$ to a large upper bound, saying $R_c$. Perfect
convergence can be achieved by using $R_c = 30\sigma$ throughout the
paper. Since integral \eqref{eqn:kernel-inte} is independent of the
density of the system, it is calculated only once and stored.

Similarly, for Eqn. \eqref{eqn:tmp3}, the Fourier mode of the the mean error
force is (in terms of Fourier transform):
\begin{align}\label{eqn:tmp17}
  \langle\Delta\v F\rangle^\wedge(\v k) =
  \hat{\v f}^c(\v k)\,
  \hat\rho(\v k),
\end{align}
where the Fourier transform of the complementary force is given by
\begin{align}\label{eqn:tmp18}
  \hat{\v f}^c(\v k) 
  & = 
  2\pi\frac{\v k}k\, i\,
  \bigg\{
  r_c^2\,u(r_c)
  \bigg[\,
  \frac{2\cos(2\pi kr)}{2\pi kr}
  -\frac{2\sin(2\pi kr)}{(2\pi kr)^2}
  \,\bigg] 
  -
  \int_{r_c}^\infty 2r\,u(r)\sin(2\pi kr)\,\d d\v r
  \bigg\}.
\end{align}

>>>>

The naive calculation of the convolutions in the error estimate \eqref{eqn:tmp5} requires
a computational cost of $\mathcal O(N^2)$. To estimate the force error
on the fly, the computational cost of the estimate should be reduced to
at least $\mathcal O(N\log N)$.  This is achieved by  using the fast Fourier transform. We
assume the simulation box is uniformly divided in to small bins.  The
positions of these bins are
\begin{align}
  \v r_{i_1, i_2, i_3} & =
  \frac{i_1}{K_1}\v a_1 + 
  \frac{i_2}{K_2}\v a_2 + 
  \frac{i_3}{K_3}\v a_3,\qquad 0\leq i_\alpha< K_\alpha
\end{align}
where $\v a_\alpha$ are box vectors, and $K_\alpha$ is the number of
division on each direction. In the reciprocal space, a corresponding
lattice is set up:
\begin{align}
  \v k_{k_1, k_2, k_3} & =
  k_1\v a_1^\ast +
  k_2\v a_2^\ast +
  k_3\v a_3^\ast, \qquad 0\leq k_\alpha<K_\alpha
\end{align}
where $\v a_\alpha^\ast$ are reciprocal box vectors defined by $\v
a_\alpha\cdot\v a_\beta^\ast = \delta_{\alpha\beta}$.  In each bin the
error and the particles number density are assumed to be constant,
then, a Fourier expansion can be obtained:
\begin{align}\nonumber
  \mathcal E^2_{\textrm{homo}}(\v r_{i_1,i_2,i_3})
  & =
  \frac1V\sum_{k_1=-\infty}^\infty\sum_{k_2=-\infty}^\infty\sum_{k_3=-\infty}^\infty
  [{\mathcal E}^2_{\textrm{homo}}]^{\wedge}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg] \\\label{eqn:tmp13}
  & =
  \frac1V\sum_{k_1=0}^{K_1}\sum_{k_2=0}^{K_2}\sum_{k_3=0}^{K_3}
  \hat{K}_{\textrm{homo}}(\v k'_{k_1,k_2,k_3})\:
  \hat{\rho}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
Here $\vert\v f^c(\v r)\vert^2$ is denoted by
$K_{\textrm{homo}}(\v r)$.
The wedge notation $\wedge$ means Fourier transform. The prime on $\v k_{k_1,k_2,k_3}$ means
that the periodic image $k_\alpha - K_\alpha$ should use instead of
$k_\alpha$ when $k_\alpha \geq K_\alpha/2$.
The second equation in \eqref{eqn:tmp13}
holds due to the fact:
$[\mathcal E^2]^\wedge = [K_{\textrm{homo}}\ast\rho]^\wedge = \hat{K}_{\textrm{homo}} \times \hat\rho$. The Fourier transform of $\rho$ and $K_{\textrm{homo}}$  are
\begin{align}
  \hat\rho(\v k_{k_1,k_2,k_3})
  & =
  \frac{V}{K_1K_2K_3}
  \sum_{i_1=0}^{K_1}\sum_{i_2=0}^{K_2}\sum_{i_2=0}^{K_2}
  \rho(\v r_{i_1,i_2,i_3})
  \exp\bigg[
  -2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg],
\end{align}
and
\begin{align}\label{eqn:f-kernel-homo}
  \hat K_{\textrm{homo}}(\v k)
  &=
  \int_{\mathbb R^3}K_{\textrm{homo}}(\v s)e^{-2\pi i\v k\cdot\v s}\,\d d\v s.
\end{align}
Most of the short-range
interactions are isotropic, so  $K_{\textrm{homo}}(\v r) =
K_{\textrm{homo}}(r)$, then the three-dimensional integral in
Eqn. \eqref{eqn:f-kernel-homo} can be simplified to a one-dimensional
integral:
\begin{align}\label{eqn:kernel-inte}
  \hat{K}_{\textrm{homo}}(\v k)
  =
  % \int_{\mathbb R^3}K_{\textrm{homo}}(r)e^{-2\pi i\v k\cdot\v r}\d d\v r
  % =
  \int_{r_c}^\infty \frac{2 s}k\, K_{\textrm{homo}}(s) \sin(2\pi ks)\,\d ds,
\end{align}
which can be calculated by numerical
integration from $r_c$ to a large upper bound, saying $R_c$. Perfect
convergence can be achieved by using $R_c = 30\sigma$ throughout the
paper. Since integral \eqref{eqn:kernel-inte} is independent of the
density of the system, it is calculated only once and stored.

Similarly, for Eqn. \eqref{eqn:tmp3}, the Fourier mode of the the mean error
force is (in terms of Fourier transform):
\begin{align}\label{eqn:tmp17}
  \langle\Delta\v F\rangle^\wedge(\v k) =
  \hat{\v f^c}(\v k)\,
  \hat\rho(\v k),
\end{align}
where the Fourier transform of the complementary force is given by
\begin{align}\label{eqn:tmp18}
  \hat{\v f^c}(\v k) 
  & = 
  2\pi\frac{\v k}k\, i\,
  \bigg\{
  r_c^2\,u(r_c)
  \bigg[\,
  \frac{2\cos(2\pi kr_c)}{2\pi kr_c}
  -\frac{2\sin(2\pi kr_c)}{(2\pi kr_c)^2}
  \,\bigg] 
  -
  \int_{r_c}^\infty 2s\,u(s)\sin(2\pi ks)\,\d ds
  \bigg\},
\end{align}
where $u(r)$ is the isotropic pairwise potential.
